Population structure and phenotypic variation of *Sclerotinia sclerotiorum* from dry bean throughout the United States, and from Mexico, France, and Australia

### 2017-06-27 doc/manuscript/manuscript.Rmd
update manuscript to be peerj style

---
title: Population structure and phenotypic variation of *Sclerotinia sclerotiorum* from dry bean throughout the United States, and from Mexico, France, and Australia
author:
  - name: Zhian N. Kamvar
    email: zkamvar@unl.edu
    affiliation: University of Nebraska-Lincoln
  - name: Bimal Sajeewa Amadarasa
    email: bamadarasa2@unl.edu
    affiliation: University of Nebraska-Lincoln
  - name: R. Jhala
    email: rjhala2@unl.edu
    affiliation: University of Nebraska-Lincoln
  - name: James Steadman
    email: jsteadman1@unl.edu
    affiliation: University of Nebraska-Lincoln
  - name: Sydney E. Everhart
    email: everhart@unl.edu
    affiliation: University of Nebraska-Lincoln
    corresponding: everhart@unl.edu
address:
  - code: University of Nebraska-Lincoln
    address: Department of Plant Pathology, University of Nebraska-Lincoln, Lincoln 68583
abstract: >
  `r paste(readLines("abstract.md"), collapse = "\n  ")`
author_summary: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit.

bibliography: mybibfile.bib
output: rticles::plos_article
csl: plos.csl
---

@@

*Sclerotinia sclerotiorum* (Lib.) de Bary is an ascomycete plant pathogen found worldwide.
This is a necrotrophic pathogen that is primarily homothallic (self-fertilization) (Bolton et al., 2006) (Sexton et al., 2006).
It has a host range of more than 400 species of plants belonging to 75 families (Boland and Hall, 1994).
Economically important crops affected by this pathogen include sunflower, soybean, oilseed rape, dry bean, chickpea, peanut, dry pea, lentils and vegetables (Bolton et al., 2006). *Sclerotinia sclerotiorum* causes white mold disease on dry bean (Phaseolus vulgaris L.), which can be a yield-limiting in temperate climates (Steadman, 1983).

@@

@@

A distance-based redundancy analysis (dbRDA) (Legendre and Anderson, 1999) was performed with the function capscale in the vegan 2.4-3 package in R (https://CRAN.R-project.org/package=vegan) to delineate relative contribution of phenotypic, spatial, and temporal components in influencing genetic composition of the populations. dbRDA is a method of conducting constrained ordinations on data matrices consisting of non-Euclidean distance measures.
RDA is a multivariate linear regression analysis of independent (explanatory: eg. environmental and phenotypic) variables on dependent (non-explanatory or response) variables.
In our case, a pairwise Bruvo's (Bruvo et al., 2004) distance matrix was generated from microsatellite data using poppr 2.4.1 (Kamvar et al., 2014; Kamvar et al., 2015) and employed it as the dependent matrix for RDA analysis.
To identify environmental variable(s) correlated with genetic variation, a forward-backward selection process was applied with the vegan function ordistep to identify the set of variables that explained more variation than predicted with the global model containing all variables.
An ANOVA was then performed to test for significance of the reduced model and marginal effects using 999 permutations.
The varpart function of vegan was used to determine variation partitioning of explanatory variables.

@@

To assess the connectivity of shared genotypes across populations, we first tabulated the multilocus genotypes shared between at least two populations (defined as states or countries) with the poppr function mlg.crosspop() (Kamvar et al. 2014).
From these data, we were able to construct a graph with populations as nodes and shared genotypes as edges (connections) between the nodes using the R packages igraph version 1.0.1 (Csardi and Nepusz 2006), dplyr version 0.5.0 (Wickham and Francois 2016), and purrr version 0.2.2 (Wickham 2016).
Each node was weighted by the fraction of shared MLGs.
Each edge represents a single MLG, but because a single MLG could cross populations, that MLG may be represented by (n*(n-1))/2 edges where n represents the number of populations it crosses.
Edges were weighted by 1- Psex, where Psex is the probability of encountering the same genotype via two independent meiotic events (Arnaud-Haond 2007, Parks and Werth 1993).
We then used the cluster_optimal() function from igraph to define communities (Csardi and Nepusz 2006).
The graph was plotted using the R packages ggplot2 version 2.2.1 (Wickham 2009) and ggraph version 1.0.0 (Pedersen 2017).
To ensure that we were able to capture the same community signal, we additionally ran this analysis including the 5 polymorphic markers described above.

@@

DAPC is a multivariate method designed to identify and describe clusters of genetically related individuals.
Discriminant analysis maximizes separation between groups while minimizing variation within groups (Jombart et al., 2010).
The find.clusters function in Adegenet 2.0.1 (Jombart, 2008) uses a k-means clustering algorithm to identify the optimum number of clusters.
It was not able to identify a single optimum and identified a range of optimal clusters between 5-15.
Consequently, DAPC was performed several times with different optimal number of clusters to determine whether isolates (N = 318) fell into each cluster have any biological meaning such as belonging to a particular host or came from the same region etc.  the best summaries for the total isolates (N = 318).
Since dbRDA identified region, host, and sampling year as the top three variables influencing microsatellite data, DAPC was also performed on data defined according to each of these variables, separately.
Isolates (N = 318) were assigned to different levels of the three factors (region, host, year) and assessed predefined group size (original number of isolates at each level of the factor), post group size (number of isolates assigned by the DAPC algorithm) and the proportion of the original isolates assigned to the original group after analysis.
This was performed to determine the suitability of explanatory variables (i.e. region, host, and year) in describing the clusters generated.
Clone-corrected isolates sampled in different years from the white mold nurseries (wmn) were tested for population differentiation over time by DAPC analysis.
Only WA, MI, and NE isolates were qualified since we filtered out the states that had at least ≥7 samples/year with at least two years of data for the analysis.
Lastly, we pooled all isolates, clone corrected them (N = 165) and performed DAPC using the function find.clusters as described above to delineate how many clusters were formed by these unique genotypes.
AMOVA was used to test the significance of clusters resulted by different DAPC analyses using GenAlEx version 6.41 (Peakall and Smouse, 2012).
PhiPT which is an analogue of Wright's FST was used to ascertain the significance of populations.
An FST value of 0.05 or less is generally considered as reasonably low and may be interpreted to mean that structuring between populations is weak (Balloux and Ugon-Moulin, 2002; Wright, 1978).

@@

Populations from different regions were compared with AMOVA, where differentiation was low among regions (PhiPT  = 0.11; P = 0.001).
Pairwise comparisons of all populations according to region showed MX and CA populations were significantly different across all populations at P ≤ 0.001.
The PhiPT values of Pairwise comparisons between CA and other populations ranged from 0.116 to 0.389 and from 0.355 to 0.493 between MX and other population.
The CA population showed evidence of sexual recombination ( = 0.008; P = 0.27, Ia = 0.08; P = 0.3) while the  value of AU was 0.12 with a high P of 0.05 (Ia = 0.70; P = 0.05), whereas all other populations showed linkage disequilibrium for loci tested indicating high possibility of clonal reproduction.

@@

We tested the clone corrected data for evidence of sexual outcrossing within *S. sclerotiorum* populations in different regions and different years/region.
CA showed evidence of sexual outcrossing (Ia = 0.08, P.Ia = 0.317,  = 0.008, P. = 0.317) while all other regions implied clonal propagation (data not shown).
Five populations showed evidence of sexual outcrossing: CA in 2004, FR in 2004 and 2005, OR in 2003, and WA in 2008.
However, after clone-correction these populations had 6-7 isolates/population (Table 7) and therefore, may not be sufficient to make inferences about reproductive mode.
Only exception was the WA_2008 population, which consisted of 12 isolates.
Within sampling locations white mold nurseries (wmn) were tested for linkage disequilibrium since same wmn fields were used for sampling over different years.
Only CA wmn showed evidence of sexual reproduction ( = 0.008; P = 0.29) while all other wmn locations suggested clonal reproduction.
The DAPC analyses of WA, and MI isolates from same wmn fields over different years are in Fig. 10.
Since *S. sclerotiorum* isolates are predominantly clonal, we wanted to explore whether there was any introduction of new isolates from other fields or if there were evidence of sexual reproduction.
The AMOVA results for different clusters of Fig. 10A revealed significance difference between WA 2008 isolates from the rest (PhiPT ≥ 0.206, P = 0.001).
The 2003 and 2005 isolates were genetically similar (PhiPT = 0.056, P = 0.129).
MI isolates from 2008 and 2003 (Fig. 10B) showed significant difference (PhiPT = 0.127, P = 0.04) while 2005 isolates did not differ molecularly from both 2008 and 2003 isolates (PhiPT ≤ 0.078, P ≥ 0.076).
NE wmn isolates sampled in 2004 and 2008 were genetically similar (PhiPT = 0.042, P = 0.179) (DAPC results not shown).
The standardized index of association of wmn isolates from each region (i.e.
MI, WA, and NE) for above mentioned sampling years suggested clonal reproduction ( ≥ 0.05, P ≤ 0.002).

@@

Our study was conducted to characterize *S. sclerotiorum* isolates from different geographic areas to determine important phenotypic variables influencing genotypes; to ascertain reproductive behavior of the pathogen; to test whether different sampling years or locations have any effect on genotype/sexuality; and to identify abundant and unique genotypes across isolates and geographies.
The dbRDA revealed region, year, and host as the most influential phenotypic variables on genotypes.
However, further molecular analysis of pathogen isolates from selected hosts did not show any significant genetic difference among populations.
According to dbRDA, virulence and location of isolates had little or no effect.
Previous studies on *S. sclerotiorum* from alfalfa soil (Attanayake et al., 2012) and potato (Atallah et al., 2004), and canola (Attanayake et al., 2013) plants reported little or no correlation of genotypes on pathogen virulence.
This is important information for resistant breeding since genotypes of *S. sclerotiorum* isolates seem to depend to a certain degree on different levels of region and also on environmental conditions of the sampling year.
For instance our analysis indicated that MX, 2009, and 2005 were highly influential predictors of *S. sclerotiorum* genotypes (Fig. 7).
Conversely, France and Australia, which are geographically distant to North America, were not within the top 10 predictors of genetic composition of isolates (Fig. 7).
According to dbRDA, Mexican region imposed the highest impact on microsatellite loci of *S. sclerotiorum* isolates.
We used Bruvo's distance for calculating genetic distances of different isolates since it incorporates stepwise mutation model and applicable to microsatellite analysis.

@@

Contrary to above results Aldrich-Wolfe et al. (2015) who did a survey on *S. sclerotiorum* from different host crops in the North Central states of the United States found a single microsatellite haplotype to be prevalent throughout the region.
A study in the UK that examined this pathogen from 12 populations in different crop plants and weed species *Ranunculus acris* (meadow buttercup) also reported one microsatellite haplotype was widely distributed across hosts and areas (Clarkson et al., 2013).
Reason for our isolates from dry bean not to show a widespread MLG may be due to extensive sexual outcrossing and movement sclerotia by human activities.
Though conditions of outcrossing for *S. sclerotiorum* is not well studied, it is believed weather and agricultural conditions play a role in production and release of ascospores of this pathogen (Huang, 1991; Sexton et al., 2006).
One study found release of spores from apothecia declined over 36- 168 H at low RH of 65-75% but not at high RH of 90-95% (Clarkson et al., 2003).
The same study reported increased ascospores discharge at 20-25° C compared to 15° C and the survival of *S. sclerotiorum* ascospores was reduced at high temperature and RH.
Our samples were collected from 2003 to 2012 and except for the dry bean nurseries, came from different fields.
Different agro-ecological conditions prevailed during sampling seasons may have resulted in different rates of sexual outcrossing of the pathogen and hence reduced the chance of extracting a common genotype.
When conditions are favorable for ascospore production, release, and survival, co-infection of a single site on a plant by multiple ascospores is possible (Sexton et al., 2006).
This may eventually result in facilitating outcrossing and recombination.

>>>>

---
title: Population structure and phenotypic variation of *Sclerotinia sclerotiorum* from dry bean throughout the United States, and from Mexico, France, and Australia
preprint: false
author:
  - name: Zhian N. Kamvar
    email: zkamvar@unl.edu
    affiliation: 1
  - name: Bimal Sajeewa Amadarasa
    email: bamadarasa2@unl.edu
    affiliation: 1
  - name: R. Jhala
    email: rjhala2@unl.edu
    affiliation: 1
  - name: James Steadman
    email: jsteadman1@unl.edu
    affiliation: 1
  - name: Sydney E. Everhart
    email: everhart@unl.edu
    affiliation: 1
    corresponding: true
affiliation:
  - code: 1
    address: Department of Plant Pathology, University of Nebraska-Lincoln, Lincoln 68583
abstract: >
  `r paste(readLines("abstract.md"), collapse = "\n  ")`
# author_summary: |
#   Lorem ipsum dolor sit amet, consectetur adipiscing elit.
header-includes: >
  \usepackage{indentfirst}
bibliography: mybibfile.bib
biblio-style: apalike
output: 
  rticles::peerj_article:
    citation_package: natbib
    base_format: bookdown::pdf_document2 # for using \@ref()
---

@@

*Sclerotinia sclerotiorum* (Lib.) de Bary is an ascomycete plant pathogen found worldwide.
This is a necrotrophic pathogen that is primarily homothallic (self-fertilization) (Bolton et al., 2006) (Sexton et al., 2006).
It has a host range of more than 400 species of plants belonging to 75 families (Boland and Hall, 1994).
Economically important crops affected by this pathogen include sunflower, soybean, oilseed rape, dry bean, chickpea, peanut, dry pea, lentils and vegetables (Bolton et al., 2006). *Sclerotinia sclerotiorum* causes white mold disease on dry bean (*Phaseolus vulgaris* L.), which can be a yield-limiting in temperate climates (Steadman, 1983).

@@

deleteme: @Dirac1953888

@@

A distance-based redundancy analysis (dbRDA) (Legendre and Anderson, 1999) was performed with the function capscale in the vegan 2.4-3 package in R (https://CRAN.R-project.org/package=vegan) to delineate relative contribution of phenotypic, spatial, and temporal components in influencing genetic composition of the populations. dbRDA is a method of conducting constrained ordinations on data matrices consisting of non-Euclidean distance measures.
RDA is a multivariate linear regression analysis of independent (explanatory: eg. environmental and phenotypic) variables on dependent (non-explanatory or response) variables.
In our case, a pairwise Bruvo's [@bruvo2004simple] distance matrix was generated from microsatellite data using poppr 2.4.1 [@kamvar2014poppr; @kamvar2015novel] and employed it as the dependent matrix for RDA analysis.
To identify environmental variable(s) correlated with genetic variation, a forward-backward selection process was applied with the vegan function ordistep to identify the set of variables that explained more variation than predicted with the global model containing all variables.
An ANOVA was then performed to test for significance of the reduced model and marginal effects using 999 permutations.
The varpart function of vegan was used to determine variation partitioning of explanatory variables.

@@

To assess the connectivity of shared genotypes across populations, we first tabulated the multilocus genotypes shared between at least two populations (defined as states or countries) with the poppr function mlg.crosspop() [@kamvar2014poppr].
From these data, we were able to construct a graph with populations as nodes and shared genotypes as edges (connections) between the nodes using the R packages igraph version 1.0.1 (Csardi and Nepusz 2006), dplyr version 0.5.0 (Wickham and Francois 2016), and purrr version 0.2.2 (Wickham 2016).
Each node was weighted by the fraction of shared MLGs.
Each edge represents a single MLG, but because a single MLG could cross populations, that MLG may be represented by (n*(n-1))/2 edges where n represents the number of populations it crosses.
Edges were weighted by 1- Psex, where Psex is the probability of encountering the same genotype via two independent meiotic events (Arnaud-Haond 2007, Parks and Werth 1993).
We then used the cluster_optimal() function from igraph to define communities (Csardi and Nepusz 2006).
The graph was plotted using the R packages ggplot2 version 2.2.1 (Wickham 2009) and ggraph version 1.0.0 (Pedersen 2017).
To ensure that we were able to capture the same community signal, we additionally ran this analysis including the 5 polymorphic markers described above.

@@

DAPC is a multivariate method designed to identify and describe clusters of genetically related individuals.
Discriminant analysis maximizes separation between groups while minimizing variation within groups (Jombart et al., 2010).
The find.clusters function in Adegenet 2.0.1 (Jombart, 2008) uses a k-means clustering algorithm to identify the optimum number of clusters.
It was not able to identify a single optimum and identified a range of optimal clusters between 5-15.
Consequently, DAPC was performed several times with different optimal number of clusters to determine whether isolates (N = 318) fell into each cluster have any biological meaning such as belonging to a particular host or came from the same region etc.  the best summaries for the total isolates (N = 318).
Since dbRDA identified region, host, and sampling year as the top three variables influencing microsatellite data, DAPC was also performed on data defined according to each of these variables, separately.
Isolates (N = 318) were assigned to different levels of the three factors (region, host, year) and assessed predefined group size (original number of isolates at each level of the factor), post group size (number of isolates assigned by the DAPC algorithm) and the proportion of the original isolates assigned to the original group after analysis.
This was performed to determine the suitability of explanatory variables (i.e. region, host, and year) in describing the clusters generated.
Clone-corrected isolates sampled in different years from the white mold nurseries (wmn) were tested for population differentiation over time by DAPC analysis.
Only WA, MI, and NE isolates were qualified since we filtered out the states that had at least $\geq$ 7 samples/year with at least two years of data for the analysis.
Lastly, we pooled all isolates, clone corrected them (N = 165) and performed DAPC using the function find.clusters as described above to delineate how many clusters were formed by these unique genotypes.
AMOVA was used to test the significance of clusters resulted by different DAPC analyses using GenAlEx version 6.41 (Peakall and Smouse, 2012).
PhiPT which is an analogue of Wright's FST was used to ascertain the significance of populations.
An FST value of 0.05 or less is generally considered as reasonably low and may be interpreted to mean that structuring between populations is weak (Balloux and Ugon-Moulin, 2002; Wright, 1978).

@@

Populations from different regions were compared with AMOVA, where differentiation was low among regions (PhiPT  = 0.11; P = 0.001).
Pairwise comparisons of all populations according to region showed MX and CA populations were significantly different across all populations at P $\leq$ 0.001.
The PhiPT values of Pairwise comparisons between CA and other populations ranged from 0.116 to 0.389 and from 0.355 to 0.493 between MX and other population.
The CA population showed evidence of sexual recombination ( = 0.008; P = 0.27, Ia = 0.08; P = 0.3) while the  value of AU was 0.12 with a high P of 0.05 (Ia = 0.70; P = 0.05), whereas all other populations showed linkage disequilibrium for loci tested indicating high possibility of clonal reproduction.

@@

We tested the clone corrected data for evidence of sexual outcrossing within *S. sclerotiorum* populations in different regions and different years/region.
CA showed evidence of sexual outcrossing (Ia = 0.08, P.Ia = 0.317,  = 0.008, P. = 0.317) while all other regions implied clonal propagation (data not shown).
Five populations showed evidence of sexual outcrossing: CA in 2004, FR in 2004 and 2005, OR in 2003, and WA in 2008.
However, after clone-correction these populations had 6-7 isolates/population (Table 7) and therefore, may not be sufficient to make inferences about reproductive mode.
Only exception was the WA_2008 population, which consisted of 12 isolates.
Within sampling locations white mold nurseries (wmn) were tested for linkage disequilibrium since same wmn fields were used for sampling over different years.
Only CA wmn showed evidence of sexual reproduction ( = 0.008; P = 0.29) while all other wmn locations suggested clonal reproduction.
The DAPC analyses of WA, and MI isolates from same wmn fields over different years are in Fig. 10.
Since *S. sclerotiorum* isolates are predominantly clonal, we wanted to explore whether there was any introduction of new isolates from other fields or if there were evidence of sexual reproduction.
The AMOVA results for different clusters of Fig. 10A revealed significance difference between WA 2008 isolates from the rest (PhiPT $\geq$ 0.206, P = 0.001).
The 2003 and 2005 isolates were genetically similar (PhiPT = 0.056, P = 0.129).
MI isolates from 2008 and 2003 (Fig. 10B) showed significant difference (PhiPT = 0.127, P = 0.04) while 2005 isolates did not differ molecularly from both 2008 and 2003 isolates (PhiPT $\leq$ 0.078, P $\geq$ 0.076).
NE wmn isolates sampled in 2004 and 2008 were genetically similar (PhiPT = 0.042, P = 0.179) (DAPC results not shown).
The standardized index of association of wmn isolates from each region (i.e.
MI, WA, and NE) for above mentioned sampling years suggested clonal reproduction ( $\geq$ 0.05, P $\leq$ 0.002).

@@

Our study was conducted to characterize *S. sclerotiorum* isolates from different geographic areas to determine important phenotypic variables influencing genotypes; to ascertain reproductive behavior of the pathogen; to test whether different sampling years or locations have any effect on genotype/sexuality; and to identify abundant and unique genotypes across isolates and geographies.
The dbRDA revealed region, year, and host as the most influential phenotypic variables on genotypes.
However, further molecular analysis of pathogen isolates from selected hosts did not show any significant genetic difference among populations.
According to dbRDA, virulence and location of isolates had little or no effect.
Previous studies on *S. sclerotiorum* from alfalfa soil (Attanayake et al., 2012) and potato (Atallah et al., 2004), and canola (Attanayake et al., 2013) plants reported little or no correlation of genotypes on pa25thogen virulence.
This is important information for resistant breeding since genotypes of *S. sclerotiorum* isolates seem to depend to a certain degree on different levels of region and also on environmental conditions of the sampling year.
For instance our analysis indicated that MX, 2009, and 2005 were highly influential predictors of *S. sclerotiorum* genotypes (Fig. 7).
Conversely, France and Australia, which are geographically distant to North America, were not within the top 10 predictors of genetic composition of isolates (Fig. 7).
According to dbRDA, Mexican region imposed the highest impact on microsatellite loci of *S. sclerotiorum* isolates.
We used Bruvo's distance for calculating genetic distances of different isolates since it incorporates stepwise mutation model and applicable to microsatellite analysis.

@@

Contrary to above results Aldrich-Wolfe et al. (2015) who did a survey on *S. sclerotiorum* from different host crops in the North Central states of the United States found a single microsatellite haplotype to be prevalent throughout the region.
A study in the UK that examined this pathogen from 12 populations in different crop plants and weed species *Ranunculus acris* (meadow buttercup) also reported one microsatellite haplotype was widely distributed across hosts and areas (Clarkson et al., 2013).
Reason for our isolates from dry bean not to show a widespread MLG may be due to extensive sexual outcrossing and movement sclerotia by human activities.
Though conditions of outcrossing for *S. sclerotiorum* is not well studied, it is believed weather and agricultural conditions play a role in production and release of ascospores of this pathogen (Huang, 1991; Sexton et al., 2006).
One study found release of spores from apothecia declined over 36- 168 H at low RH of 65-75% but not at high RH of 90-95% (Clarkson et al., 2003).
The same study reported increased ascospores discharge at 20-25$^{\circ}$ C compared to 15$^{\circ}$ C and the survival of *S. sclerotiorum* ascospores was reduced at high temperature and RH.
Our samples were collected from 2003 to 2012 and except for the dry bean nurseries, came from different fields.
Different agro-ecological conditions prevailed during sampling seasons may have resulted in different rates of sexual outcrossing of the pathogen and hence reduced the chance of extracting a common genotype.
When conditions are favorable for ascospore production, release, and survival, co-infection of a single site on a plant by multiple ascospores is possible (Sexton et al., 2006).
This may eventually result in facilitating outcrossing and recombination.
